Mathematical Description

### 2023-06-15 100-math-description/GR-sslcm-math-description.Rmd

---
title: "Mathematical Description"
subtitle: "State-Space Life Cycle  Model for Grande Ronde Spring Chinook Salmon"
output: 
  bookdown::pdf_document2:
    keep_tex: true
    toc: true
    toc_depth: 3
    number_sections: true
    citation_package: natbib
  # bookdown::word_document2
fontfamily: mathpazo
geometry: [top=1in,bottom=1in,left=0.25in,right=1.75in]
linkcolor: cyan
urlcolor: cyan
citecolor: black
bibliography: "GR-sslcm-cites.bib"
link-citations: "`r ifelse(knitr::is_latex_output(), 'true', 'false')`"
biblio-style: apalike
csl: "https://raw.githubusercontent.com/citation-style-language/styles/master/canadian-journal-of-fisheries-and-aquatic-sciences.csl"
editor_options: 
  chunk_output_type: console
header-includes:
  - \usepackage{microtype}
  - \usepackage{float}
  - \usepackage{xcolor}
  - \definecolor{my-blue}{HTML}{E6F0FF}
  - \usepackage[textwidth=1.5in]{todonotes}
  - \setuptodonotes{backgroundcolor=my-blue,textcolor=black,linecolor=ProcessBlue,bordercolor=my-blue,size=footnotesize}
  - \usepackage[labelfont={bf,sc},labelsep=period]{caption}
  - \captionsetup{width=\textwidth}
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \cfoot{\thepage}
  - \renewcommand{\headrulewidth}{0.5pt}
  - \renewcommand{\footrulewidth}{0.5pt}
  - \lfoot{State-Space Life Cycle Model\\Mathematical Description}
  - \rfoot{Grande Ronde Spring Chinook Salmon}
  - \usepackage{titlesec}
  - \titleformat{\section}{\bfseries\LARGE\scshape}{\thesection}{1em}{}
  - \titleformat{\subsection}{\bfseries\large}{\thesubsection}{1em}{}
  - \titleformat{\subsubsection}{\itshape}{\thesubsubsection}{1em}{}
  - \renewcommand{\contentsname}{Table of Contents}
  - \renewcommand{\eqref}[1]{\ref{#1}}
  - \newcommand{\fall}{\mathrm{fall}}
  - \newcommand{\spring}{\mathrm{spring}}
  - \newcommand{\NOR}{\mathrm{NOR}}
  - \newcommand{\HOR}{\mathrm{HOR}}
  - \newcommand{\CAT}{\mathrm{CAT}}
  - \newcommand{\LOS}{\mathrm{LOS}}
  - \newcommand{\MIN}{\mathrm{MIN}}
  - \newcommand{\UGR}{\mathrm{UGR}}
  - \newcommand{\WUL}{\mathrm{WUL}}
---

@@

The purpose of the state-space life cycle model for Grande Ronde spring Chinook salmon is to synthesize data sets that have been collected as part of routine monitoring programs into a self-contained and internally consistent population dynamics model.
Such a synthesis can provide richer insights about unobserved quantities (e.g., total parr recruitment abundance) and the relationships between them than allowed by more piecemeal analyses.
Further, because the model estimates expected values and the magnitude of inter-annual variability in a time series framework, the output will be useful for structuring simulation models of future population dynamics to evaluate different management strategies.
Although this is the ultimate goal and purpose of developing the state-space life cycle model described herein, we do not discuss this concept further, as it is beyond the scope of the statistical population dynamics model, which is the topic of this document.

The state-space life cycle model tracks the fates of multiple cohorts of fish from egg to adulthood within a statistical population dynamics model.
The model receives input data in the form of empirical observations or pre-processed output from other models (e.g., to derive survival rates from PIT tag records) and estimates unknown parameters that must be internally consistent with the data and one another.
The model is considered "state-space" because it captures two discrete types of variability: (_i_) process noise, i.e., variability in the biological and environmental processes that produce population dynamics and outcomes and (_ii_) observation error, i.e., variability in the measurement processes that result in the observed data representing an imperfect perception of the true value being measured.
As such, the model is inherently a time series model: spawners in one year produce eggs, which move through several life stages and ultimately return as spawners in future years -- it is the same pool of fish at one life stage that is decremented (via survival rate transition probabilities) to become a pool of fish used in the calculations for a later stage of the life cycle.
We have described the life cycle model calculations as divided into three distinct life "phases" (Figure \@ref(fig:diagram)): freshwater juvenile, ocean juvenile, and freshwater adult.

The model has several dimensions and the majority of processes are assumed to operate differently (i.e., have varying numerical values) along each dimension. These include:

* **Age of Return** -- Anadromous Chinook salmon in the Grande Ronde basin return predominately as age-3, age-4, or age-5 adults. This means that monitoring activities in one year of adult returns sample fish that are progeny from from 3 consecutive brood (i.e., i.e., spawning) years, creating time linkages in the time series that are important to account for.
* **Origin** -- Adults returning in any given year may be the result of either natural spawning/rearing in the wild or via captive spawning/rearing in a hatchery setting. These are critical to track separately because the productivity and capacity of naturally spawning salmon are key outputs of the model.
* **Juvenile Migratory Strategy** -- Grande Ronde spring Chinook salmon juveniles exhibit two main phenological migratory strategies which may have different vital rates and require additional complexity in the observation model.
* **Population** -- Four tributaries of the Grande Ronde River have been monitored using largely consistent methods for nearly three decades; these include Catherine Creek, Lostine River, Minam River, and Upper Grande Ronde River. We track the abundance and survival of each separately because they are monitored separately and local environmental conditions and/or genetic adaptations may result in population dynamics differing in terms of productivity, capacity, or overall life cycle fitness.

We used the Bayesian mode of inference to fit the model to data, as it is intuitive for the necessary integration over the many layers of uncertainty (i.e., process vs. observational variability).
Further, Bayesian inference provides an intuitive framework to constrain parameters to take on only values known _a priori_ to be plausible -- we intentionally employ nearly exclusively uninformative prior information, but in several cases we were required to use auxiliary information in the form of reasonably strong priors (Table \@ref(tab:param-table)).
An additional benefit of using the Bayesian framework is the seamless propagation of uncertainty from estimated parameters to derived quantities (of which there are many) -- all unknown quantities receive a posterior distribution.
We employed Markov Chain Monte Carlo methods (implemented via the JAGS probabilistic programming language) to sample from the joint posterior distribution of all unknown quantities in the model.

@@

We have striven for consistency in our mathematical representation of quantities in the model but this has created somewhat non-intuitive conventions which require devoted explanation.
All quantities are defined (i.e., symbology and the specific equations that use them) in Tables \@ref(tab:index-table), \@ref(tab:symbol-table), \@ref(tab:param-table), and \@ref(tab:data-table) for reference and a catalog of all equations is provided in Table \@ref(tab:equation-table).

* For simplicity, we use the term "adult" to refer to any mature individual returning to spawn including age-3 "jacks".
* Life stage-specific states are defined by capital letters, e.g., parr ($P$), smolt ($M$), or spawners ($S$)
* Parameters (e.g., survival terms, non-survival transition probabilities, coefficients in relationships, etc.) are denoted by Greek symbols, e.g., $\phi$ for survival rates or $\psi$ for maturation rates.
* Subscripts denote the index of a quantity, e.g., $y$ denotes a specific year and $j$ a specific population (see Table \@ref(tab:index-table)).
* We use the convention $b$ to indicate "before event" and $a$ to indicate "after event" for a quantity that occurs at the same life stage. E.g., $M^{b}$ represents smolt abundance _before_ out-of-basin migration and $M^{a}$ represents smolt abundance _after_ out-of-basin migration. As such, superscripts are most often used syntactically, not mathematically. E.g., $M^{b}$ is not "$M$ raised to the power $b$"; if we must raise the quantity $M^{b}$ to the power $c$, it will be written $(M^{b})^c$.
* Survival terms to transition from state $A$ to state $B$ in year $y$ for population $j$ are denoted by $\phi^{A \to B}_{y,j}$.
* We denote the expected value of a stochastic process with a dot, e.g., $\dot{\phi}^{A \to B}_j$ -- this is the value expected in the absence of process variability.
* In some cases, we specify calculations performed on vectors rather than scalars (i.e., a scalar being a single element from a vector or higher-dimensional array). Vectors are denoted via boldface, and the dimension of the vector is denoted using 1:$n$ syntax. E.g., $\boldsymbol{\phi}^{A \to B}_{y,1:n_j}$ represents a vector of survival terms from state $A$ to state $B$ in year $y$, where each element of the vector represents the value for each population.
* Mathematical operations performed on two vectors $\boldsymbol{C}$ and $\boldsymbol{D}$ are performed in an element-wise fashion, and the resulting vector will have identical dimensions to both $\boldsymbol{C}$ and $\boldsymbol{D}$.
* In some cases, a calculation is performed differently for a subset of the possible index values. For example, to denote that $k$ should take on only the values of 1 and 3 (not 2), we would write, $k \in [1,3]$.
* $\log\left(x\right)$ represents the natural logarithm of $x$ and it is implied that the constraint $x > 0$ is satisfied.
* $\logit\left(p\right)$ represents the log odds of $p$ (i.e., $\log\left[p/(1-p)\right])$) and it is implied that the constraint $0 < p < 1$ is satisfied.
* All quantities that have been empirically observed (or externally estimated from empirically observed data, and supplied to the model as "data") and are used in a likelihood component are denoted by a "hat", e.g., a survival term ($\hat{\phi}^{A \to B}_{y,j}$) and its logit-normal standard error ($\hat{\sigma}_{\phi^{A \to B}_{y,j}}$, which we use as a direct estimate of observation error variability).
* Stochastic processes are represented by $x_{y} \sim \mathcal{F}(\theta_{1}, \theta_{2})$, where $x_{y}$ is a random variable, $\mathcal{F}$ is some probability density (or mass) function ($\mathcal{N}$ is univariate normal, $\mathcal{MVN}$ is multivariate normal, $\mathcal{B}$ is binomial, and $\mathcal{M}$ is multinomial), and $\theta_{1}$ and $\theta_{2}$ are parameters of the probability distribution; generally $\theta_{1}$ is a location parameter specifying the expected value of $x_{y}$ and $\theta_{2}$ is a dispersion parameter specifying the variability of $x_{y}$ around the expected value.

Process noise is introduced by assuming year-specific values are multivariate logit-normal random variables around the deterministic value.
This is done to enable modeling covariability in process noise among populations (or origins, in the few cases where population structure is not modeled).
The covariance matrix of these random variables are denoted by $\Sigma$, and have a common structure for all model components.
Take the hypothetical survival term $\phi^{A \to B}_{y,j}$ as an example.
We denote its random variability across years and populations as:

@@

In cases where a quantity varies annually and is different by origin but not by population, $\Sigma$ would have dimensions $n_o \times n_o$:

@@

We estimated these covariance matrices by assigning their inverse ($\Sigma^{-1}$) a scaled Wishart prior distribution (Table \@ref(tab:param-table)).
Although the stochastic process requires the $\Sigma$ term, it is the component $\sigma$ and $\rho$ terms that we were interested in for inference.

@@

We begin our description of the life cycle at the egg stage immediately after spawning.
We assumed total egg production was the sum product of age-specific spawner abundance (from eq. \@ref(eq:get-Sa), below), proportion female, and fecundity:

@@

We assumed that spawners age-4 or age-5 were 50% female and age-3 spawners were all male (i.e., $\Omega_{k=1} = 0;\;\Omega_{k \in [2,3]} = 0.5$; approximately equal to values estimated from carcass data) and estimated fecundity ($f_{y,k,j}$) from length-fecundity relationships based on Grande Ronde-origin hatchery broodstock.
We supplied both quantities to the model assuming no error or inter-annual variability.

We assumed expected egg-to-parr survival was density-dependent following a Beverton-Holt relationship with productivity parameter $\alpha_j$ and capacity parameter $\beta_j$:

@@

To facilitate later analyses (not documented here) investigating the effects of habitat restoration and climate change on population dynamics, we modeled parr capacity as a function of weighted usable habitat length ($\WUL_j$)^[Derivation of $\WUL_j$ is too detailed to describe fully here.
Briefly, the relative ability of stream segments with different habitat characteristics to hold rearing parr was expressed in terms of the density of fish predicted to reside there by a complex generalized linear mixed model.
The model was fitted to density estimates based on detectability-adjusted [@staton-etal-2022] summer snorkel counts with local habitat characteristics (e.g., pool frequency, water temperature, large wood density, etc.) as predictor variables.
The weight of each stream segment was assigned based on the predicted density relative value to the segment with the highest value; $\WUL_j$ was then equal to the sum of segment-specific weight and length within the known spawning and rearing extent for each population.].
We modeled parr recruitment capacity ($\beta_j$) as a function of this $\WUL_j$ metric:

@@

We modeled realized (i.e., with process noise) egg-to-parr survival as a multivariate logit-normal random variable around the expected value ($\dot{\phi}^{E \to P^{b}}_{y,j}$) from eq. \@ref(eq:get-phi-E-Pb-dot).
Rather than assume that inter-annual variability in egg-to-parr survival beyond that explained by density-dependence is completely random across years, we allowed the model to account for serial autocorrelation in egg survival anomalies via a lag-1 autoregressive process [AR(1); coefficient denoted $\kappa^{E \to P^{b}}_{j}$], wherein the white noise portion of the anomaly has covariance matrix $\Sigma_{\phi^{E \to P^{b}}}$:

@@

Parr recruitment ($P^{b}_{y,j}$) to the end of summer was then the product of total egg production and egg-to-parr survival:

@@

In addition to density-dependent egg-to-parr survival (eq. \@ref(eq:get-phi-E-Pb-dot)), we also wished to model the size of parr at the end of summer as a function of density.
It is well known that juvenile survival is related to body size \todo{Insert references about survival being related to size} and if parr size is density-dependent, that would provide the possibility of modeling variability in survival later in life (overwinter rearing and migration out-of-basin) based on a density-dependent process that occurred earlier.
We modeled expected parr mean length ($\dot{L}^{P^{b}}_{y,j}$, mm fork length) as a power function of egg density, which expressed on the logarithmic scale was:

@@

Like many populations of stream-type spring Chinook salmon in the Columbia River basin \todo{Insert references about other CRB populations displaying the fall/spring migratory strategy}, there is life history diversity among the juveniles with respect to their migratory phenology and habitat types occupied for rearing overwinter.
Some portion of the summer parr ($P^{b}_{y,j}$)  migrate out of the headwaters near where they were spawned in the fall and rear farther downstream (termed "fall migrants" and indexed by $i = \fall$) and the remaining portion migrate out the following spring (termed "spring migrants" and indexed by $i = \spring$).
Due to this difference in migratory phenology, these two groups of fish are monitored separately \todo{Insert a reference to a recent ODFW screw trap monitoring report where it is stated (or implied) that migratory strategies are monitored separately.} depending on when they pass the rotary screw trap located in each tributary\todo{Create a map and include reference it here}.
To enable building an observation model that could accommodate this feature of the populations in our model, we apportioned summer parr to each of these two strategies:

@@

where $P^{a}_{y,i,j}$ is migratory strategy-specific parr abundance and $\pi_{y,i,j}$ is the proportion that take on each migratory strategy.
We assumed the expected value of this proportion for fall migrants ($\dot{\pi}_{i = \fall,j}$) was time-constant and modeled the realized values as having a multivariate logit-normal distribution with covariance matrix $\Sigma_{\pi_{i = \fall}}$:

@@

Since there are only two migratory strategies, we obtained the proportion that are spring migrants as the complement ($\pi_{y,i = \spring,j} = 1 - \pi_{y,i = \fall,j}$).

@@

We modeled expected survival from the parr (age-1) to in-basin smolt (age-2) stage (termed "overwinter" survival and denoted by $\dot{\phi}^{P^{a} \to M^{b}}_{y,i,j}$) as a logit-linear function of parr mean length:

@@

where $L^{*P^{b}}_{y,j}$ is parr mean length scaled and centered by subtracting the historical average observed parr mean length for population $j$ from $L^{P^{b}}_{y,j}$ and dividing by the historical observed standard deviation in mean length across years.
Then, for each migratory strategy separately, overwinter survival was a multivariate logit-normal random variable around the expected value with covariance matrix $\Sigma_{\phi^{P^{a} \to M^{b}}}$ (assumed common among migratory strategies):

@@

Since hatchery-origin smolts are released into the population in spring (see Section \@ref(hatchery-inputs)), we introduce the origin index ($o$) starting at the in-basin smolt stage prior to out-migration ($M^{b}_{y,i,o,j}$).
The abundance of natural-origin (indexed by $o=\NOR$) smolt was then:

@@

We further wished to use the mean length of these surviving smolt ($L^{M^{b}}_{y,j}$) to later explain variability in survival along the migration out-of-basin.
Because only spring migrants are measured for smolt mean length in the spring (as they pass the screw trap), we were forced to assume equal lengths for fall migrant smolt and we thus omit the $i$ subscript from these quantities related to smolt mean length.
Preliminary analyses on the observed mean length data revealed that smolt mean length was positively related with parr mean length, but that the relationship was non-linear.
To capture this non-linearity, we modeled the expected multiplicative change in mean length between the parr and smolt stage as a log-linear function of (standardized) parr mean length:

@@

and obtained the realized multiplicative change in mean length as being multivariate log-normally distributed around $\dot{\Delta}^{L^{P^{b}} \to L^{M^{b}}}_{y,j}$ with covariance matrix $\Sigma_{\Delta^{L^{P^{b}} \to L^{M^{b}}}}$:

@@

Note the summation across migratory strategies ($i$) in eq. \@ref(eq:get-O0), which indicates that the fates of fish with these different strategies were assumed to be the result of identical processes following arrival to LGR.

@@

for the summer tagging group,

@@

for the fall tagging group,

@@

for the spring tagging group, and

@@

which we used as the multinomial expected frequency to build the likelihood:

>>>>

---
title: "Mathematical Description"
subtitle: "State-Space Life Cycle  Model for Grande Ronde Spring Chinook Salmon"
output: 
  bookdown::pdf_document2:
    keep_tex: true
    toc: true
    toc_depth: 3
    number_sections: true
    citation_package: natbib
  # bookdown::word_document2:
  #   reference_docx: "C:/Users/bstaton/Desktop/Staton/3_general/Rmd-docx-template/Rmd-docx-template-basic-TNR.docx"
  #   number_sections: true
fontfamily: mathpazo
geometry: [top=1in,bottom=1in,left=1in,right=1in]
linkcolor: cyan
urlcolor: cyan
citecolor: black
bibliography: "GR-sslcm-cites.bib"
link-citations: "`r ifelse(knitr::is_latex_output(), 'true', 'false')`"
biblio-style: apalike
csl: "https://raw.githubusercontent.com/citation-style-language/styles/master/canadian-journal-of-fisheries-and-aquatic-sciences.csl"
editor_options: 
  chunk_output_type: console
header-includes:
  - \usepackage{natbib}
  - \usepackage{soul}
  - \setcitestyle{aysep={}}
  - \usepackage[hang,flushmargin,bottom]{footmisc}
  - \usepackage{microtype}
  - \usepackage{float}
  - \usepackage{xcolor}
  - \usepackage[textwidth=1.5in,tickmarkheight=3pt]{todonotes}
  - \setuptodonotes{backgroundcolor=black!40,textcolor=white,linecolor=black!65,bordercolor=black!65,size=footnotesize}
  - \usepackage[labelfont={bf,sc},labelsep=period]{caption}
  - \captionsetup{width=\textwidth}
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \cfoot{\thepage}
  - \renewcommand{\headrulewidth}{0.5pt}
  - \renewcommand{\footrulewidth}{0.5pt}
  - \lfoot{State-Space Life Cycle Model\\Mathematical Description}
  - \rfoot{Grande Ronde Spring Chinook Salmon}
  - \usepackage{titlesec}
  - \titleformat{\section}{\bfseries\LARGE\scshape}{\thesection}{1em}{}
  - \titleformat{\subsection}{\bfseries\large}{\thesubsection}{1em}{}
  - \titleformat{\subsubsection}{\itshape}{\thesubsubsection}{1em}{}
  - \renewcommand{\contentsname}{Table of Contents}
  - \renewcommand{\eqref}[1]{\ref{#1}}
  - \newcommand{\fall}{\mathrm{fall}}
  - \newcommand{\spring}{\mathrm{spring}}
  - \newcommand{\NOR}{\mathrm{NOR}}
  - \newcommand{\HOR}{\mathrm{HOR}}
  - \newcommand{\CAT}{\mathrm{CAT}}
  - \newcommand{\LOS}{\mathrm{LOS}}
  - \newcommand{\MIN}{\mathrm{MIN}}
  - \newcommand{\UGR}{\mathrm{UGR}}
  - \newcommand{\WUL}{\mathrm{WUL}}
---

@@

The purpose of developing the state-space life cycle model for Grande Ronde spring Chinook salmon is to estimate posterior distributions of population dynamics parameters (e.g., average survival, rearing capacity, coefficients of relationships linking life stages, etc.).
These posterior distributions can ultimately be sampled from to parameterize simulation models that evaluate habitat restoration and climate change scenarios grounded in empirical monitoring data.
The model tracks the cohort-specific abundance of fish through various life stages and links these "state variables" over time across cohorts via production relationships (e.g., spawners produce eggs) or within cohorts via transition probabilities (e.g., parr become smolt after surviving overwinter).
The model integrates many different data sources and fits to them simultaneously in a single joint likelihood; these data sources include rotary screw trap passage estimates, a variety of PIT tag-derived survival estimates, mean length data, hatchery smolt releases, adult return abundance, and the composition of adult returns by age and origin; other data sources that are used but are not fitted to include hatchery smolt release abundances, harvest information, fecundity estiamtes, and numbers of returning adults removed for hatchery operations.
Four spawning populations within the Grande Ronde basin are simultaneously analyzed: Catherine Creek (CAT), Lostine River (LOS), Minam River (MIN) and the upper Grande Ronde River (UGR; `r highlight("FIGURE REFERENCE: map")`), and where relevant, the model stratifies life stages by juvenile migratory strategy (fall vs. spring migrants), rearing origin type (spawned in hatchery vs. natural setting), and adult return age (total age of 3, 4, or 5 years).

Central to the notion of a "state-space" model is the partitioning of variability in the observed data sets into two types: (_i_) process noise, i.e., variability in the biological and environmental processes that produce population outcomes and (_ii_) observation error, i.e., variability in the measurement processes that result in an imperfect perception of the true value being measured.
This is accomplished by (_i_) specifying a set of process equations that represent expected population responses to intrinsic (e.g., density affects survival) and extrinsic (e.g., quantity of available habitat affects capacity) drivers, (_ii_) adding process noise to obtain the "realized" or "latent" (i.e., true but partially or imperfectly observed) state, and (_iii_) assuming the data values have been observed conditioned on the latent values.
The model does not attempt to estimate the magnitude of both process and observation noise variance; it is supplied estimates of the latter and attributes the remaining noise unexplained by process equations to process noise.
This aspect critical because only variability due to the processes is of interest for parameterizing simulation models of future population dynamics, yet the data sets vary in their precision.
Further, the state-space model is inherently a time series model, making it suitable for handling the time lags and linkages involved with modeling age-structured populations.

Bayesian methods were used to fit the state-space life cycle model to data, as it is intuitive for the necessary integration over the many layers of uncertainty (i.e., process vs. observational variability).
Further, Bayesian inference provides an intuitive framework to constrain parameters to take on only values known _a priori_ to be plausible -- mostly uninformative priors were intentionally used but several cases involved priors that were moderately or strongly informative.
An additional benefit of using the Bayesian framework is the seamless propagation of uncertainty from estimated parameters to derived quantities -- all unknown quantities receive a posterior distribution.
Markov Chain Monte Carlo methods [MCMC, implemented via the JAGS v4.3.1 probabilistic programming language, @plummer-2003] were used to sample from the joint posterior distribution.

@@

The state-space life cycle model symbology aims for consistency in mathematical representation of quantities but this has created some conventions that require devoted explanation.
All quantities are defined in Tables \@ref(tab:index-table), \@ref(tab:symbol-table), \@ref(tab:param-table), and \@ref(tab:data-table) and a catalog of all equations is provided in Table \@ref(tab:equation-table).

* For simplicity, the term "adult" is used to refer to any mature individual returning to spawn, including age-3 "jacks".
* Life stage-specific states are defined by capital letters, these include: eggs ($E$), parr ($P$), smolt ($M$), ocean juveniles ($O$), adults returning to river ($R$), or adults with potential to spawn ($S$).
* Parameters (e.g., survival terms, non-survival transition probabilities, coefficients in relationships, etc.) are predominately denoted by Greek symbols and are used consistently where possible, e.g., $\phi$ for survival rates or $\psi$ for maturation rates.
* Subscripts denote placement of a quantity in a larger array (i.e., the index), e.g., $y$ denotes a specific year and $j$ a specific population (see Table \@ref(tab:index-table)).
* Superscripts are most often used syntactically to further describe the quantity, not in a mathematical sense.
E.g., $M^{b}$ represents smolt abundance _before_ out-of-basin migration and $M^{a}$ represents smolt abundance _after_ out-of-basin migration.
As such, $M^{b}$ is not "$M$ raised to the power $b$"; if the quantity $M^{b}$ must be raised to the power $c$, it will be written $(M^{b})^c$.
* Survival terms to transition from state $A$ to state $B$ in year $y$ for population $j$ are denoted by $\phi^{A \to B}_{y,j}$.
* The expected value of a stochastic process is expressed by a dot, e.g., $\dot{\phi}^{A \to B}_j$ -- this is the value expected in the absence of process (biological/environmental) variability.
* In some cases, calculations are performed on vectors rather than scalars.
Vectors are denoted via boldface, and the dimension of the vector is denoted using 1:$n$ syntax. E.g., $\boldsymbol{\phi}^{A \to B}_{y,1:n_j}$ represents a vector of parameters representing the transition (survival) probability from state $A$ to state $B$ in year $y$, where each element stores the value for each population.
* Mathematical operations performed on two vectors $\boldsymbol{C}$ and $\boldsymbol{D}$ are performed in an element-wise fashion, and the resulting vector will have identical dimensions to both $\boldsymbol{C}$ and $\boldsymbol{D}$.
* In some cases, a calculation is performed differently for a subset of the possible index values. For example, to denote that $k$ should take on only the values of 1 and 3 (not 2), we would write, $k \in [1,3]$.
* $\log\left(x\right)$ represents the natural logarithm of $x$ and it is implied that the constraint $x > 0$ is satisfied.
* $\logit\left(p\right)$ represents the log odds of $p$ (i.e., $\log\left[p/(1-p)\right])$) and it is implied that the constraint $0 < p < 1$ is satisfied.
* All quantities are used in a likelihood component are denoted by a "hat", e.g., a survival term ($\hat{\phi}^{A \to B}_{y,j}$) and its logit-normal standard error ($\hat{\sigma}_{\phi^{A \to B}_{y,j}}$, which is used as a direct estimate of observation error variability).
* Stochastic processes are represented by $x_{y} \sim \mathcal{F}(\theta_{1}, \theta_{2})$, where $x_{y}$ is a random variable, $\mathcal{F}$ is some probability density (mass) function ($\mathcal{N}$ is univariate normal, $\mathcal{MVN}$ is multivariate normal, $\mathcal{B}$ is binomial, and $\mathcal{M}$ is multinomial), and $\theta_{1}$ and $\theta_{2}$ are parameters of the probability distribution.

Process noise is modeled by assuming year-specific values are multivariate logit-normal random variables around the deterministic value.
This enables modeling covariability in demographic rates [@riecke-etal-2019; @bouchard-etal-2022] among populations or origins, which is advantageous for estimation involving incompletely overlapping time series and quantifies synchrony in demographic rates which influences the stability of the return aggregated across populations [@thorson-etal-2014; @connors-etal-2022].
The covariance matrices of these random variables are denoted by $\Sigma$, and have a common structure for all model components.
Take the hypothetical survival term $\phi^{A \to B}_{y,j}$ as an example.
The random process introducing inter-annual variability and inter-population covariability is expressed as:

@@

In cases where a quantity varies annually and covaries by origin but is identical across populations, $\Sigma_{\phi^{A \to B}}$ would have dimensions $n_o \times n_o$:

@@

These covariance matrices were estimated by assigning their inverse ($\Sigma^{-1}$) a scaled Wishart prior distribution [@gelman-etal-2014; @plummer-2017, Table \@ref(tab:param-table)].
Although the stochastic process requires the $\Sigma$ term, it is the component $\sigma$ and $\rho$ terms that were of interest for inference.

@@

The life cycle begins at the egg stage immediately after spawning.
Total egg production was the sum product of age-specific spawner abundance ($S^{a}_{y,k,o,j}$, from eq. \@ref(eq:get-Sa), below), proportion female-at-age ($\Omega_k$), and fecundity-at-age ($f_{y,k,j}$):

@@

Spawners of age-4 or age-5 were assumed to be 50% female and age-3 spawners were assumed 100% male (i.e., $\Omega_{k=1} = 0;\;\Omega_{k \in [2,3]} = 0.5$; approximately equal to values estimated from carcass data).
Fecundity was predicted from length-fecundity relationships based on Grande Ronde-origin hatchery broodstock and time-varying mean length-at-age data.
Both $\Omega_k$ and $f_{y,k,j}$ were assumed known without error.

Expected egg-to-parr survival was assumed to be a density-dependent process following Beverton-Holt dynamics with "productivity" parameter $\alpha_j$ (i.e., theoretical maximum egg-to-parr survival probability, in the absence of density effects) and capacity parameter $\beta_j$ (i.e., theoretical maximum parr recruitment abundance):

@@

To facilitate later analyses (not documented here) investigating the effects of habitat restoration and climate change on population dynamics, parr capacity was modeled as a function of weighted usable habitat length specific to each population ($\WUL_j$)^[Derivation of $\WUL_j$ is too detailed to describe fully here.
Briefly, the relative ability of stream segments with different habitat characteristics to hold rearing parr was expressed in terms of the density of fish predicted to reside there by a complex generalized linear mixed model.
The model was fitted to density estimates based on detectability-adjusted [@staton-etal-2022] summer snorkel counts with local habitat characteristics (e.g., pool frequency, water temperature, large wood density, etc.) as predictor variables.
The weight of each stream segment was assigned based on the predicted density relative value to the segment with the highest value; $\WUL_j$ was then equal to the sum of segment-specific weight and length within the known spawning and rearing extent for each population.]:

@@

Realized (i.e., with process noise) egg-to-parr survival was a multivariate logit-normal random variable around the expected value ($\dot{\phi}^{E \to P^{b}}_{y,j}$) from eq. \@ref(eq:get-phi-E-Pb-dot).
Rather than assume egg-to-parr survival anomalies (i.e., inter-annual variability in egg-to-parr survival beyond that explained by density-dependence) are completely random across years, a lag-1 autoregressive process [AR(1); coefficient denoted $\kappa^{E \to P^{b}}_{j}$] was used to account for serial autocorrelation:

@@

Thus, the covariance matrix $\Sigma_{\phi^{E \to P^{b}}}$ captures variability in the white noise (i.e., non-correlated) portion of the survival anomaly.
Total parr recruitment ($P^{b}_{y,j}$) was then the product of total egg production and egg-to-parr survival:

@@

In addition to density-dependent egg-to-parr survival (eq. \@ref(eq:get-phi-E-Pb-dot)), there was reason to expect that parr growth/size is density-dependent, both from the literature on salmonid early life history [@grant-imre-2005; @copeland-venditti-2009; @walters-etal-2013; @myrvold-kennedy-2015; @grossman-simon-2020] and previous analyses on Grande Ronde Spring Chinook salmon populations [@cooney-etal-2017; @justice-etal-2023].
Given the established relationships between size and survival  [@zabel-achord-2004; @hostetter-etal-2015], growth and density [e.g., @grant-imre-2005], and survival and density [@achord-etal-2003; @walters-etal-2013], it is possible that growth could be a useful mechanism to model density effects on post-recruitment juvenile survival.
Expected parr mean length ($\dot{L}^{P^{b}}_{y,j}$, mm fork length) was modeled as a power function of egg density, which expressed on the logarithmic scale was:

@@

Like many populations of stream-type spring Chinook salmon in the Columbia River basin, those in the Grande Ronde display life history diversity with respect to juvenile migratory phenology and overwinter rearing habitat use [copeland-etal-2014].
Some portion of parr recruits ($P^{b}_{y,j}$)  migrate from the headwaters rearing areas in the fall and rear overwinter farther downstream (termed "fall migrants" and indexed by $i = \fall$) and the remaining portion migrate the following spring during the out-of-basin migration (termed "spring migrants" and indexed by $i = \spring$).
Due to this difference in migratory phenology, these two groups of fish are monitored separately (`r highlight("REPORT CITATION NEEDED")`) depending on when they pass the rotary screw trap located in each tributary (`r highlight("FIGURE REFERENCE: map")`).
Separate abundance and survival data in the observation model required that the process model track fish using these strategies separately, thus they were apportioned:

@@

where $P^{a}_{y,i,j}$ is migratory strategy-specific parr abundance and $\pi_{y,i,j}$ is the proportion that take on each migratory strategy.
The expected value of the proportion of fall migrants ($\dot{\pi}_{i = \fall,j}$) was assumed to be time-constant and the realized values were modeled as having a multivariate logit-normal distribution with covariance matrix $\Sigma_{\pi_{i = \fall}}$:

@@

With only two migratory strategies, the proportion that were spring migrants was obtained as the complement: $\pi_{y,i = \spring,j} = 1 - \pi_{y,i = \fall,j}$.

@@

The model assumed that some inter-annual variability in overwinter survival (i.e., the transition from age-1 parr to age-2 smolt prior to out-migration) can be explained by parr mean length^[The assumption that parr length is related to overwinter survival can be relaxed by forcing $\gamma_{1,j}$ in eq. \@ref(eq:get-phi-Pa-Mb-dot) to zero rather than assigning it an uninformative prior distribution; in this case overwinter survival would be assumed to fluctuate around a time-constant expected value.].
This was achieved by modeling expected overwinter survival (denoted by $\dot{\phi}^{P^{a} \to M^{b}}_{y,i,j}$) as a logit-linear function of parr mean length:

@@

where $L^{*P^{b}}_{y,j}$ is parr mean length centered and scaled based on the mean and standard deviation of observed mean lengths (denoted $\hat{L}^{P^{b}}_{y,j}$) across years for population $j$, respectively.
Thus, $\gamma_{0,i,j}$ represents the expected log-odds of survival for type $i$ in population $j$ in a year with average parr mean length, and $\gamma_{1,j}$ represents the expected change in the log-odds for every additional standard deviation increase in parr mean length.
Then, for each migratory strategy separately, realized overwinter survival was a multivariate logit-normal random variable around the expected value with covariance matrix $\Sigma_{\phi^{P^{a} \to M^{b}}}$ (assumed common among migratory strategies):

@@

The abundance of natural origin smolt (indexed by $o=\NOR$, hatchery-origin smolt releases described in Section \@ref(hatchery-inputs)) before out-migration was then:

@@

Similar to the use of parr mean length to explain inter-annual variability in overwinter survival, smolt mean length ($L^{M^{b}}_{y,j}$) may be useful for explaining inter-annual variability in out-migration survival.
Only spring migrant smolt have been measured for mean length (as they pass the screw trap), thus the $i$ subscript is omitted from quantities related to smolt mean length.
Preliminary analyses revealed that smolt mean length was positively related to parr mean length, but that the relationship was non-linear.

Preliminary analyses on the observed mean length data revealed that smolt mean length was positively related with parr mean length, but that the relationship was non-linear [@justice-etal-2023].
This non-linearity was captured by modeling the expected multiplicative change in mean length from the parr to smolt stages as a log-linear function of (scaled and centered) parr mean length:

@@

The realized multiplicative change in mean length was multivariate log-normally distributed around $\dot{\Delta}^{L^{P^{b}} \to L^{M^{b}}}_{y,j}$ with covariance matrix $\Sigma_{\Delta^{L^{P^{b}} \to L^{M^{b}}}}$:

@@

The summation across migration strategies ($i$) in eq. \@ref(eq:get-O0) indicates that the fates of fish with differing migratory strategies were not tracked separately after the $M^{a}_{y,i,o,j}$ stage.

@@

Fall tag group:

@@

Winter tag group:

@@

Hatchery smolt tag group:

@@

which was used as the multinomial expected frequency to build the likelihood:
